Density-dependent selection and the limits of relative fitness

### 2017-12-08 manuscript.tex
l

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, $s$ will often depend on density. Eq.~\eqref{eq:canonical} is then no longer a complete description of selection (we would also need to specify a model for density). Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated compared to the constant-$s$ case. Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.

@@

It is widely recognized that the relative fitness description of selection widely used by evolutionists is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}. As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, unfairly invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

Having defended Eq.~\eqref{eq:canonical}

, relative fitness does indeed break down if strong selection occurs in a way that changes (adult) population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}). In our lottery model, the this can only occur if types differ in $c$ as well as $b$ or $d$. The reason for this is that density-dependent ($c$) and density-determining traits ($b$ and $d$) are completely separate. This may seem odd given the behavior of most models in the density-dependent literature, so it worth investigating this point further.

The essential distinction between selection in uncrowded and crowded conditions is the absence or presence of interactions between types (these interactions may be indirect, as in the case of consumable resource competition). That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: (i) some types expand more rapidly in the absence of interactions between types or (ii) some types are superior in their interactions with other types. This distinction captures a key property of crowding, namely that you must displace others to grow.

By contrast, the separation of traits in the $r$/$K$ scheme is presented as a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density (e.g. \citep{gill_1974}). As we have seen, MacArthur's general argument \citep{macarthur_1967} does not justify this dichotomy, because selection in crowded populations also involves inter-type competition represented by the Lotka-Volterra $\alpha_{ij}$ coefficients; selection on the latter has been termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. In the Lotka-Volterra model, $r_i$ belongs to (i) and the inter-type coefficients $\alpha_{ij}$ ($i\neq j$) belong to (ii). But $\alpha_{ii}$, the inverse of type $i$'s carrying capacity $K_{ii}$, straddles both (i) and (ii). For this reason, $K$-selection --- selection for greater carrying capacity --- is a particularly confusing way to characterize selection in crowded populations.

In our lottery model, $b$ and $d$ are interaction-independent traits, whereas $c$ is entirely about interactions between types.

(factors which cause density trajectories to move ``outwards'' in Fig.~\ref{fig:Ksel}a)

(factors which cause density trajectories to move diagonally in Fig.~\ref{fig:Ksel}b).

\citep{chesson_2000}

Another common feature of density-regulated growth models is the presence of traits which both regulate density and are subject to density-dependent selection, such as the $\delta$ trait in the simple birth-death model Eq.~\eqref{eq:simplebirthdeath}. The $b$/$\delta$ parameterization in Eq.~\eqref{eq:simplebirthdeath} emphasizes the obvious point that traits which control density need not be subject to density-dependent selection. The $K$ parameter in the standard $r/K$ parameterization of the logistic model has been particularly misleading in this respect, because it is solely responsible for the regulation of density and is subject to density-dependent selection. Nevertheless, $\delta$-like traits arise automatically in the simplest linear models of density-regulated growth, including the logistic and Lotka-Volterra models, regardless of how they are parameterized, and are thus deeply ingrained in the literature on density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012}. In our lottery model, $b$ and $d$ control density, but do not exhibit density-dependent selection, and thus do not pose the same threat to Eq.~\eqref{eq:canonical} as a $\delta$-like trait.

It is not obvious that we should expect this as a general rule. Much of our intuition about type-specific regulation of density (and density-dependent selection in general)  Consumable resource models are a better guide, because they explicitly model the processes underlying an important form of competition. Density regulation is type-specific in well-mixed resource competition models, but for many populations consumable resources are not well-mixed. Spatial localization of consumable resources (e.g. due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and is subsumed into the competitive ability $c$. Density regulation will then apply at the population level similarly to the lottery model, allowing $N$ to change via $b$ or $d$ without affecting Eq.~\eqref{eq:canonical}.

Relative fitness models truly breaks down if $N$ is far from equilibrium and selection is density-dependent. For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness. Our density-dependent lottery model may not be a particularly good description of \textit{Drosophila} ecology, but the close connection between our model and Wright-Fisher is particularly useful, because drift in our model should behave broadly similarly. Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.

>>>>

The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations. Since crowded and uncrowded conditions are so different, $s$ will often depend on density. Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for density. Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated compared to the constant-$s$ case. Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant. The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.

@@

It is widely recognized that the relative fitness description of selection widely used by evolutionists is poorly integrated with the ecological literature on population growth \citep{mallet_2012}. This is not simply a lack of ecological realism on the part of relative fitness models. In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}. As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated. In our lottery model, this excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.

Having highlighted this redeeming quality of Eq.~\eqref{eq:canonical}, we now turn to its breakdown. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent. In our lottery model, this can only occur if types differ in $c$ as well as $b$ or $d$. Why does our model need joint trait variation when the classical density-dependent selection literature is built on traits that behave like the $\delta$ in Eq.~\eqref{eq:canonical} \citep{christiansen_2004}? To briefly review: Based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Thus we are left with $r$-selection (uncrowded) versus $K$- or $\alpha$-selection (crowded). The latter two forms of selection in crowded populations behave like selection on $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).

By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent. This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1976}. Both are needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium). In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.

Relative fitness models truly break down if $N$ is far from equilibrium and selection is density-dependent. For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness. Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly. Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.
